---
title: "Introduction"
output:
  html_document:
    toc: true
---

Hello! Please click open the "dtavir-workshop.Rproj" in the folder you downloaded before the workshop. Then, in the right pane of RStudio that this opened, click the file called "01-introuction.Rmd". If that worked, you are all set for now!

# Welcome

- About you
- About me

## About This Workshop

- Why a workshop?
    - Help you get past initial barriers.
    - Misplaced comma shouldnt keep you from learning R.
    - Provide foundation for going further on your own.
- No coding experience required.
- Will learn to transform and visualize data in R.
- Will use higher education data (IPEDS).
- Live coding, not slides.

## Writing Code

- This is an RMarkdown document.
- Code (gray areas) and regular text are mixed (white areas).
- The gray areas, called code chunks, can be run by clicking the green arrows on top left of chunk.

## Code Together: Load Packages

- Click the green arrow in the code chunk below to load packages.

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(tidyverse)
library(vized)

```

- Great, you just loaded two R packages we'll be using. More on what a package is in the next section.
- Another way to run code is to highlight the specific line(s) of code you want to run and press Ctrl + Enter.

## Code Together: Run code below with Ctrl + Enter


```{r math}

4 + 2

```

## Your Turn: Update the points.

- I'd like you to:
    1. Run the code in the gray box below and see what it does.
    2. Change the color of the points and rerun the code.
    3. Alpha can range from 0 to 1. Experiment with changing it and rerunning the code. What does alpha do?

```{r, base_chart}

ggplot(data = ipeds_coordinates) +
  geom_point(aes(x = longitude, y = latitude), 
             color = "green", 
             alpha = .9) 

```

- Things will get more difficult, but it doesn't have to be that intimidating!

## Structure of this Workshop

- The idea is to be challenged.
    - Benefits of testing (even if you don't know the answer)
- Three main types of participation throughout workshop:
    - Code Together
    - Your Turn
    - Pop quizzes
- Look for the headings for participation type.
- There is a solutions section.

## Brief Introduction to R

### Why R (or, Why Code)?

- Unconstrained by software.
- You create text:
    - Can be reused.
    - Can be read by humans.
- Scaleable.
- No copying and pasting from one software to another.
- It's free!

### R vs. RStudio

- R is a free, open-source language.
- RStudio is a company that develops open source source and sells products that facilitate its use.
- The IDE you are using is RStudio.
- RStudio's four panes.

### RMarkdown

- Mix code and text.
- Code is in gray boxes, regular text in white.
- Insert a code chunk (i.e., gray boxes).
- Notice the format and the back quotes, and what happens if that changes.
- Results appear below the code chunk.
- Naming chunks.
- Comments.
- Why RMarkdown?

### Your Turn: Knit your document

- What happens when you "Knit to HTML"? (see button at top of RStudio)

### Code Together: R Basics

- Values
    - numeric
    - character

```{r values}

```

- Objects
    1. A name without quotes.
    2. <- 
    3. Value, data, etc.

```{r objects}



```

- Functions
    1. A name without quotes
    2. Parentheses
    3. Function arguments
- Let's try the `round()` function.

```{r functions}


```

### Pop Quiz: Which one of these is a number?

- 1    
- "1"    
- "one"    
- one

### Pop Quiz: Which on of these will work?

Suppose: one <- 1

- log(1)    
- log("1")    
- log("one")    
- log(one)

### Recognizing Problems

- + sign in console
- Red x's in RStudio
- Error messages

```{r errors, error = TRUE}

round(6.4
      
nchar("supercalifragilisticexpialidocious)

"1" + "2"


```


### R packages and the tidyverse (https://www.tidyverse.org/)

- Many functions come with R when you download it (e.g., `round()`, `log()`)
- "An R package is a collection of functions, data, and documentation that extends the capabilities of base R." - R for Data Science
- To use a package, you have to do the following:
    - `install.packages("pacakgename")`: this only has to be done once.
    - `library(pacakgename)`: every time you open a new session of R and want to use package.
- This workshop focuses ont the tidyverse package, which is a collection of packages, including:
    - ggplot2
    - dplyr
    - ...
- Rather than install and load separately, you can use: 
    - `install.packages("tidyverse")`
    - `library(tidyverse)`
- Why the tidyverse?
- Many ways to get to same result!

## Conclusion

- Many ways to get same result. This workshop focuses on tidyverse.
- The answer to your question is out there:StackOverflow, blogs, message boards, free books, etc.
- Struggling is not only OK, it's encouraged!

## Your Turn: Navigate to next section

- Navigate to 02-making-a-scatterplot.Rmd
