---
title: "Making A Scatterplot"
output:
  html_document:
    toc: true
---

# Scatterplots

## Getting Started

### Code Together: Load Package & View Data

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(vized)

```

- About the vized R package and other data packages.
- In this section, we will be exploring the `ipeds_funnel` data set.
- Let's view the data set, of which there are many ways of doing so.
    - just print it.
    - `View()`
    - `glimpse()`


```{r view_data}


```

For four-year public and private-not-for profit schools, the data set contains the following columns:
    - unit_id
    - institution_name
    - sector_description
    - state_description
    - ivy_plus_exchange
    - applicants_total
    - admitted_total
    - enrolled_total
    - admit_rate
    - yield_rate
    
### Code Together: applicants_total vs. yield_rate

- The tidyverse package for visualizing data is ggplot2.
- First chart: visualize `applicants_total` vs. `yield_rate`
- What do you expect the association to be?


```{r base_chart}


```

- Troubleshooting if it didn't work
    - Balancing your parentheses
    - `+` on same line
    - misspellings

- Before addressing many problems of this chart, let's focus on the code itself.
This code contains the most basic elements you will have in a chart, namely:
  1. Initialize plot with `ggplot()` and tell data your plotting, in this case `ipeds_funnel`. 
  2. How the elements of your data are **aes**thically mapped to the chart. 
  3. The geom you want to include, which describes the type of plot.
  4. Layers of the chart are added with the `+` sign.

- The following template forms the base of the charts you will make in ggplot:

```{r template, eval = FALSE, echo = TRUE}

ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))

```

- This forms the base chart, and you can continue adding things to it, with new elements separated by a the `+` sign

### Code Together: Update the X-Axis to Log Scale

- What's wrong with the original chart and what features of it make it hard to detect any trends?

```{r log_scale}


```

### Pop Quiz: How Would You Log Transform the Y-Axis?

- Any guesses as to how we would transform the y-axis to a log scale? Try it out and see if it works.
- Going forward, it makes the most sense to leave just the axis on log scale, so let's do that.

## Finding Patterns in Data

### Pop Quiz: Schools in Upper Right

- General negative association, but a few schools on the upper right.
- Do you have a hypothesis about those schools?
    - View your data set to determine what variables could test your hypothesis.
    - BONUS: What visual elements could we add to test your hypothesis?
- Let's copy and paste our code from above to try it out.

```{r color_aes}


```

- Aesthetics you can add might include:
    - `color`
    - `size`
    - `shape`
    
### Your Turn: Try different aesthetics
    
-  Play around with the code below and add different variables to `color`, `size`, and `shape`. What do you notice? What happens with discrete vs. continuous variables? What happens when you add more than one aesthetic at a time?

```{r aes}

ggplot(data = ipeds_funnel) +
  geom_point(mapping = aes(x = applicants_total, y = yield_rate, 
                           color = ivy_plus_exchange)) +
  scale_x_log10()

```

### Code Together: Facet Your Data

- Aesthetics are not the only way to help detect patterns. 
- Using the facet functions, you can create small multiples.
- Let's do that with `sector_description` to see how IvyPlus schools are distributed across public and private institutions.
- Copy and paste code from above.


```{r facet_wrap}


```

### Your Turn: facet_wrap()

- Repeat the chart above, except this time facet your chart by `ivy_plus_exhange` instead of `sector_description`.

```{r}


```


### Code Together: facet_grid() 

- `facet_grid()` which creates a matrix of rows and columns of your data.
- It is best used when you have two discrete variables.
- Let's try it with `sector_description` and `ivy_plus_exchange`.
- Copy and paste code from above from chart we just created.


```{r facet_grid}

```

- Let's make one more adjustment before moving on.
- Many of the points overlap. Do you remember how to add transparency to the points?
- Copy and paste code from above from chart we just created.

```{r alpha}



```

- As a bonus, why does does `alpha` go outside of `aes()` instead of inside of it?

## Code Together: Multiple geoms and Local vs. Global

- Back to original chart, where we plotted `applicants_total` against `yield_rate`. 
- Seems to be a negative trend, but how do we formalize that?
- How about another `geom`? Let's try `geom_smooth`

```{r multiple_geoms}


```

- Confirms negative trend.
- Gets pulled up by elite institutions.
- Looking at the code that creates this chart, notice the redundancy?
- What happens if we put everything inside of `ggplot()` instead?

```{r local_global}


```

- So why ever use local aesthetics? 
- You may want to supply different or data aesthetics depending on the geom. Let's try an example with the chart above.

## Your Turn: Make a Scatterplot With Multiple Geoms

- Using the `ipeds_funnel` data, make a scatterplot with `geom_point()` that has the following features:
    - `applicants_total` on the x-axis.
    - `admitted_total` on the y-axis.
    - the colors of the points are represented by `sector_description`.
    - the chart contains a layer for `geom_smooth()`.
    
```{r}


```


## Conclusion

- If nothing else, remember this template.

```{r template2, eval = FALSE, echo = TRUE}

ggplot(<DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))

```

- We will return to building charts, but first let's learn how to transform data.

## Your Turn: Navigate to next section

- Navigate to 03-data-transformation.Rmd
