---
title: "Data Transformation"
output:
  html_document:
    toc: true
---

# Data Transformation

- Why start with a scatterplot? 
- Most other charts require transformation. For example, calculate:
    - means
    - sums
    - min / max
    - standard deviations
    - ...
- ggplot (can) do some data transformation, but better to do it separately.
- The tidyverse package for data transformation is dplyr.

- dplyr has five primary verbs:
    - `select()`: Pick variables by their names.
    - `arrange()`: Reorder rows.
    - `filter()`: Pick observations by their values.
    - `mutate()`: Create new variables.
    - `summarize()`: Collapse many values down to a single summary.

- 1st argument of each verb is dataframe, second is what to do with it.

## Code Together: Load Packages & View Data

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(vized)

```


- Let's take a look at the data.

```{r}

```

- Dataset for this section contains SAT scores for first-time in college students. Columns include:
    - unit_id
    - institution_name
    - percentile (25th or 75th)
    - sat_ebrw
    - sat_math

## select()

### Code Together: select()

- Let's start with `select()`
- We give it the data set, and the columns we want included.

```{r}

```

### Your Turn: select()

- Use select() to limit the `ipeds_sat` dataset to the `percentile` and `sat_math` columns.

```{r}

select(ipeds_sat, percentile, sat_math)

```

- `select()` contains many helper functions.
- Let's check out a few of them.
    - `:`
    - `contains()`
    - `starts_with()`

```{r}


```

### Pop Quiz: select()

- Which column(s) do you think the code below selects?

```{r}

select(ipeds_sat, ends_with("e"))

```

## arrange()

### Code Together: arrange()

- `arrange()` sorts data.
- Again, the first argument is the data frame, followed by columns you want sorted.
- Let's sort from lowest to highest `sat_ebrw`.

```{r}


```

- Now highest to lowest.

```{r}


```

- Now `arrange()` by multiple columns (`sat_math` & `sat_ebrw`).

```{r}


```

### Your Turn: arrange()

- Now I'd like you to arrange by `percentile` and `sat_math` (highest to lowest), `sat_ebrw` (highest to lowest).
- Which school is on top?

```{r}


```

## filter()

### Code Togther: filter()

- Pick observations (rows) by their values.
- For example, where `sat_math` is greater than 700.
- First argument is data, next are logical test(s).
- Logical tests in R include:
    - x < y	Less than
    - x > y	Greater than
    - x == y Equal to
    - x <= y Less than or equal to
    - x >= y Greater than or equal to
    - x != y Not equal to
- Let's try some out

```{r}



```

- Remember, it works with characters too! But don't forget quotes!

```{r}


```

### Pop Quiz: filter()

- `filter()` returns observations where logical criteria are true
- What do you think the following returns?

```{r}

filter(ipeds_sat, institution_name == "Florida State University")

```

- How about this one?

```{r}

filter(ipeds_sat, sat_ebrw < 500)

```

### Your Turn: filter() On One Criterion

- Now I'd like you to use filter to find all of the rows where `sat_math` is 800.

```{r}


```

### Code Together: filter() Mulitple Logical Criteria

- What about multiple logical criteria?
- For example, `sat_ebrw` is greater than 600 AND `sat_math` is greater than 600?
- Boolean oeprators include: 
    - & (and)
    - | (or)
    - ! (not)
    - () (group tests)
- Let's filter for 75th percentile scores for Florida State University.
  

```{r}


```



### Your Turn: filter() On Multiple Logical Criteria

- Please `filter()` for rows where `percentile` is 25th, `sat_ebrw` is greater than or equal to 720, and `sat_math` is greater than or equal to 720. 
- How many schools meet all of these criteria?

```{r}

```


## summarize()

### Code Together: summarize()

- `summarize()` computes tables of summaries.
- Answer questions such as "What is the...":
    - mean of x.
    - standard deviation of x.
    - frequency of x.
- Let's find the mean and standard deviation of `sat_ebrw`.
    
```{r}


```

### Your Turn: summarize()

- Using `summarize()`, instead of finding mean values with the `mean()` function, I'd like you to use the `min()` and `max` functions to find the minimum and maximum values for `sat_math`.

```{r}


```

## mutate()

### Code Together: mutate()

- `mutate()` adds new variables to your dataset.
- Let's do the following: 
    - Make a new column that is the sum of  of `sat_ebrw` and `sat_math`.
    - Make a new column that is the same as `institution_name` but all lowercase.

```{r}

mutate(ipeds_sat, 
       sat_total = sat_ebrw + sat_math,
       name_lower = tolower(institution_name))

```


### Your Turn: mutate()

- Using `mutate`,  I'd like you to create a  new column that computes the average of `sat_ebrw` and `sat_math` (i.e., `sat_ebrw` plus `sat_math` divided by 2. 
- Hint: When dividing by 2, careful with your parentheses!
    
```{r}

mutate(ipeds_sat,
       sat_mean = (sat_math + sat_ebrw) / 2) 

```

## What's Next?

- What if you wanted create a new column with `mutate()` that is the average of `sat_math` and `sat_ebrw`, `arrange()` the results in descending order,  and `select()` relevant ones?
- We can try two days based on what you already know, but in the next section we are going to learn a better way.
    - Nesting
    - Saving as objects
- Let's talk about what's wrong with each.

```{r}


```

## Your Turn: Navigate to Next Section

- Navigate to 04-pipe-operator.Rmd
