---
title: "Data Transformation"
output:
  html_document:
    toc: true
---

# Data Transformation

- Why start with a scatterplot? 
- Most other charts require transformation. For example, calculate:
    - means
    - sums
    - min / max
    - standard deviations
    - ...
- ggplot (can) do some data transformation, but better to do it separately.
- The tidyverse package for data transformation is dplyr.

- dplyr has five primary verbs:
    - `select()`: Pick variables by their names.
    - `arrange()`: Reorder rows.
    - `filter()`: Pick observations by their values.
    - `mutate()`: Create new variables.
    - `summarize()`: Collapse many values down to a single summary.

- 1st argument of each verb is dataframe, second is what to do with it.

## Code Together: Load Packages & View Data

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(vized)

```


- Let's take a look at the data.

```{r}

ipeds_sat

```

- Dataset for this section contains SAT scores for first-time in college students. Columns include:
    - unit_id
    - institution_name
    - percentile (25th or 75th)
    - sat_ebrw
    - sat_math

## select()

### Code Together: select()

- Let's start with `select()`
- We give it the data set, and the columns we want included.

```{r}

select(ipeds_sat, unit_id, institution_name)

select(ipeds_sat, unit_id, percentile, sat_ebrw)
```

### Your Turn: select()

- Use select() to limit the `ipeds_sat` dataset to the `percentile` and `sat_math` columns.

```{r}

select(ipeds_sat, percentile, sat_math)

```

- `select()` contains many helper functions.
- Let's check out a few of them.

```{r}

select(ipeds_sat, institution_name:sat_ebrw) 

select(ipeds_sat, contains("institution")) 

select(ipeds_sat, starts_with("sat"))

```

### Pop Quiz: select()

- Which column(s) do you think the code below selects?

```{r}

select(ipeds_sat, ends_with("e"))

```

## arrange()

### Code Togeher: arrange()

- `arrange()` sorts data.
- Again, the first argument is the data frame, followed by columns you want sorted.
- Let's sort from lowest to highest `sat_ebrw`.

```{r}

arrange(ipeds_sat, sat_ebrw)

```

- Now highest to lowest.

```{r}

arrange(ipeds_sat, desc(sat_ebrw))

```

- Now `arrange()` by multiple columns.

```{r}

arrange(ipeds_sat, sat_math, sat_ebrw)

```

### Your Turn: arrange()

- Now I'd like you to arrange by `percentile` and `sat_math` (highest to lowest), `sat_ebrw` (highest to lowest).
- Which school is on top?

```{r}

arrange(ipeds_sat, percentile, desc(sat_math), desc(sat_ebrw))

```

## filter()

### Code Togther: filter()

- Pick observations (rows) by their values.
- For example, where `sat_math` is greater than 700.
- First argument is data, next are logical test(s).
- Logical tests in R include:
    - x < y	Less than
    - x > y	Greater than
    - x == y Equal to
    - x <= y Less than or equal to
    - x >= y Greater than or equal to
    - x != y Not equal to
- Let's try some out

```{r}

1 > 5

2 < 11

x <- 9

x == 9


```

- Remember, it works with characters too! But don't forget quotes!

```{r}

"apple" == "apple"

"apple" == "banana"

```

### Pop Quiz: filter()

- `filter()` returns observations where logical criteria are true
- What do you think the following returns?

```{r}

filter(ipeds_sat, institution_name == "Florida State University")

```

- How about this one?

```{r}

filter(ipeds_sat, sat_ebrw < 500)

```

### Your Turn: filter() On One Criterion

- Now I'd like you to use filter to find all of the rows where `sat_math` is 800.

```{r}

filter(ipeds_sat, sat_math == 800)

```

### Code Together: filter() Mulitple Logical Criteria

- What about multiple logical criteria?
- For example, `sat_ebrw` is greater than 600 AND `sat_math` is greater than 600?
- Boolean oeprators include: 
    - & (and)
    - | (or)
    - ! (not)
    - () (group tests)
  

```{r}

filter(ipeds_sat, institution_name == "Florida State University" & percentile == "75th")

```



### Your Turn: filter() On Multiple Logical Criteria

- Please `filter()` for rows where `percentile` is 25th, `sat_ebrw` is greater than or equal to 720, and `sat_math` is greater than or equal to 720. 
- How many schools meet all of these criteria?

```{r}
filter(ipeds_sat, percentile == "25th" & sat_ebrw >= 720 & sat_math >= 720)
```


## summarize()

### Code Together: summarize()

- `summarize()` computes tables of summaries.
- Answer questions such as "What is the...":
    - mean of x.
    - standard deviation of x.
    - frequency of x.
- Let's find the mean and standard deviation of `sat_ebrw`.
    
```{r}
summarize(ipeds_sat,
          mean_sat_ebrw = mean(sat_ebrw, na.rm = TRUE),
          sd_sat_ebrw = sd(sat_ebrw, na.rm = TRUE)
)

```

### Your Turn: summarize()

- Using `summarize()`, instead of finding mean values with the `mean()` function, I'd like you to use the `min()` and `max` functions to find the minimum and maximum values for `sat_math`.

```{r}
summarize(ipeds_sat,
          min_sat = min(sat_math, na.rm = TRUE),
          max_sat = max(sat_math, na.rm = TRUE)
)
```

## mutate()

### Code Together: mutate()

- `mutate()` adds new variables to your dataset.
- Let's do the following: 
    - Make a new column that is the sum of  of `sat_ebrw` and `sat_math`.
    - Make a new column that is the same as `institution_name` but all lowercase.

```{r}

mutate(ipeds_sat, 
       sat_total = sat_ebrw + sat_math,
       name_lower = tolower(institution_name))

```


### Your Turn: mutate()

- Using `mutate`,  I'd like you to create a  new column that computes the average of `sat_ebrw` and `sat_math`. Hint: When dividing by 2, careful with your parentheses!
    
```{r}

mutate(ipeds_sat,
       sat_mean = (sat_math + sat_ebrw) / 2) 

```

## What's Next?

- What if you wanted create a new column with `mutate()` that is the average of `sat_math` and `sat_ebrw`, `arrange()` the results in descending order,  and `select()` relevant ones?
- We can try two days based on what you already know, but in the next section we are going to learn a better way.
- Let's talk about what's wrong with each.

```{r}

# first way
select(arrange(mutate(ipeds_sat,
      sat_mean = (sat_math + sat_ebrw) / 2),
desc(sat_mean)),
institution_name, percentile, sat_mean)

# second way
my_data <- mutate(ipeds_sat,
                  sat_mean = (sat_math + sat_ebrw) / 2) 

my_data <- arrange(my_data, desc(sat_mean))

select(my_data, institution_name, percentile, sat_mean)





```

## Your Turn: Navigate to Next Section
