---
title: "Grouping Your Data"
output:
  html_document:
    toc: true
---


# group_by()

## Code Together: Load Packages & View ipeds_distance

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(vized)

```

- In this section we're going to explore the `ipeds_distance` data set.

```{r, eval = FALSE}


```

- Brief Aside: When you aren't sure what a function does, use `?`. It works for data sets that are part of packages also. 
- Let's view the help page for `ipeds_distance`.

```{r}



```

- Based on what we've learned, there are lots of things we can do with this data. For example, calculate total headcount.

```{r}

```

## Code Together: Introducing group_by()

- But what if want to know the totals by `distance_type` or `level`, or `distance_type` AND `level`?
- That's where `group_by()` comes in.
- Let's try it by `distance_type`.

```{r}

```

- `group_by()` splits the results by the columns that you group by.
- How about `level` and `distance_type`?
- `group_by()` can be used with any of the five `dplyr` verbs we learned (other than `select`).
- Let's use `mutate()` to include a column with total by `level`.

## Your Turn: group_by() & summarize()

- Now I'd like you to take the `ipeds_distance` data set and combine with the pipe, `group_by()`, and `summarize()` to find the mean `headcount` by `level` and `distance_type`.
- What is mean headcount for graduate, all_distance students?

```{r}

```

## Your Turn: Fill in the Blanks.

- Now I'd like you to do the following on the `ipeds_distance` data set:
    1. Use `group_by()` to group the data by `year`, `level`, and `distance_type`.
    2. Use `summarize()` to find the sum of `headcount`.
    3. Use `filter()` to get only the rows where `year` is equal to `2018`.
    4. Use `group_by()` again to group the data by `level`.

Don't forget to remove the underscores!

Looking at the code and the result, what proportion of undergraduate students took all distance education courses in 2018?

```{r, eval = FALSE}

ipeds_distance %>% 
  group_by(year, level, _______) %>% 
  summarize(total_headcount = sum(______)) %>% 
  filter(year == ______) %>% 
  group_by(_______) %>% 
  mutate(total_by_level = sum(total_headcount),
         prop = total_headcount / total_by_level) %>% 
  ungroup() 
```

- Get in the habit of using `ungroup()`

## Your Turn: Navigate to next section

- Navigate to 06-data-import-aside.Rmd