---
title: "Graphics for Communication"
output:
  html_document:
    toc: true
---

# Improving the Look of Your Charts

- Exploring data vs. communicating data.

## Code Together: Load Packages & View Data


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(vized)
library(scales)
library(ggthemes)

```

- In this section, we will be using the `ipeds_salary` data.

```{r, eval = FALSE}

ipeds_salary %>% 
  View()

```


## Your Turn: Summarize and Plot a Bar Chart

- First, I'd like you to create the following chart from the `ipeds_salary` dataset:
    - Use `mutate()` to create a new column that finds divides `salary` by `headcount`.
    - Use `summarize()` to find the mean of your new column, grouped by `sector_description` (Hint: There is a function called `mean()`.
    - Plot the results as a bar chart (using `geom_col()`), with the mean salary you calculated on the x-axis, and `sector_description` on the y-axis.
    - Does anything about the way the results are plotted surprise you?

```{r}

ipeds_salary %>% 
  mutate(mean_salary = salary / headcount) %>% 
  group_by(sector_description) %>% 
  summarize(mean = mean(mean_salary)) %>% 
  ungroup() %>% 
  ggplot(aes(x = mean, y = sector_description)) +
  geom_col()


```

## Your Turn: Add Labels

- Below is the code from the chart we just created, but notice how there is a new layer at the end for `labs()`.
- This layer adds labels to the charts. I filled in the x label, but now I'd like you to complete the reamining ones. 
- Hint: Run the code before filling in the labels so you know what the chart looks like.
- Your answer probably won't be the same as everyone else's.
- What does each argument to `labs()`  do?

```{r}

ipeds_salary %>% 
  mutate(mean_salary = salary / headcount) %>% 
  group_by(sector_description) %>% 
  summarize(mean = mean(mean_salary)) %>% 
  ungroup() %>% 
  ggplot(aes(x = mean, y = sector_description)) +
  geom_col() +
  labs(x = "Mean Salary", y = "   ", 
       title = "  ",
       subtitle = "  ",
       caption = "  "
       )


```

## Your Turn: Add a Theme

- With chart we just created, let me show you how to add a theme.
- First, copy and paste the code for the chart we just created below.
- Here's how you add a theme:

```{r}

ipeds_salary %>% 
  mutate(mean_salary = salary / headcount) %>% 
  group_by(sector_description) %>% 
  summarize(mean = mean(mean_salary)) %>% 
  ungroup() %>% 
  ggplot(aes(x = mean, y = sector_description)) +
  geom_col() + 
  labs(x = "Mean Salary", y = "Sector", 
       title = "Average Employee Salary by Sector",
       subtitle = "Full-Time Nonmedical/Noninstructional",
       caption = "Source: IPEDS Data Center"
       ) +
  theme_minimal()

```

- Now I'd like to try and some different themes and choose one you like.

## Code Together: Reordering the Bars

- Why are the bars in the order they are?
- Reorder the levels based on mean salary.

```{r}

ipeds_salary %>% 
  mutate(mean_salary = salary / headcount) %>% 
  group_by(sector_description) %>% 
  summarize(mean = mean(mean_salary)) %>% 
  ungroup() %>% 
  mutate(sector_description = fct_reorder(sector_description, mean)) %>% 
  ggplot(aes(x = mean, y = sector_description)) +
  geom_col() + 
  labs(x = "Mean Salary", y = "Sector", 
       title = "Average Employee Salary by Sector",
       subtitle = "Full-Time Nonmedical/Noninstructional",
       caption = "Source: IPEDS Data Center"
       ) +
  theme_minimal()

```

## Code Together: Change the Bar Color

- Copy and paste code from above here.
- Regular, english colors or hex colors.
- Why is `fill` outside of `aes()` instead of inside of `aes()`?

```{r}

ipeds_salary %>% 
  mutate(mean_salary = salary / headcount) %>% 
  group_by(sector_description) %>% 
  summarize(mean = mean(mean_salary)) %>% 
  ungroup() %>% 
  mutate(sector_description = fct_reorder(sector_description, mean)) %>% 
  ggplot(aes(x = mean, y = sector_description)) +
  geom_col(fill = "blue") + 
  labs(x = "Mean Salary", y = "Sector", 
       title = "Average Employee Salary by Sector",
       subtitle = "Full-Time Nonmedical/Noninstructional",
       caption = "Source: IPEDS Data Center"
       ) +
  theme_minimal()

```

## Code Together: Changing the Scale

- Use the scales package to change the y-axis to $.

```{r}

ipeds_salary %>% 
  mutate(mean_salary = salary / headcount) %>% 
  group_by(sector_description) %>% 
  summarize(mean = mean(mean_salary)) %>% 
  ungroup() %>% 
  mutate(sector_description = fct_reorder(sector_description, mean)) %>% 
  ggplot(aes(x = mean, y = sector_description)) +
  geom_col(fill = "blue") + 
  labs(x = "Mean Salary", y = "Sector", 
       title = "Average Employee Salary by Sector",
       subtitle = "Full-Time Nonmedical/Noninstructional",
       caption = "Source: IPEDS Data Center"
       ) +
  theme_minimal() +
  scale_x_continuous(label = dollar)

```

- There is no limit to what you can do!

## Your Turn: Navigate to Next Section

- Navigate to 09-your-turn.Rmd


